A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-07-16 Manuscript/PNAS-Adams-rticle.tex
11

\author[a]{Dean C. Adams}
\author[a,b]{Erica K. Baken}
\author[b]{Michael L. Collyer}

  \affil[a]{Department of Ecology, Evolution, and Organismal Biology, Iowa State
University, Ames, Iowa, USA.}
  \affil[b]{Department of Science, Chatham University, Pittsburgh, Pennsylvania,
USA.}

@@

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Evolutionary biologists wish to quantify and compare the strength of
phylogenetic signal across traits, but analytical tools for these
comparisons are generally lacking. Here we develop a standardized effect
size based on \(\kappa\) (\(Z_\kappa\)), which measures the strength of
phylogenetic signal on a common statistical scale, and provides a
mechanism for formally comparing the strength of phylogenetic signal
across datasets. Additionally, we find that a commonly used parameter
(Pagel's \(\lambda\)) is insuitable for this purpose. Our procedure
enables biologists to quantitatively address hypotheses that compare the
strength of phylogenetic signal between various phenotypic traits, even
when those traits are found in different evolutionary lineages or have
different units or scales.}

@@

\authordeclaration{The authors declare no conflict of interest.}

@@

\begin{abstract}
Macroevolutionary studies frequently characterize the phylogenetic
signal in phenotypes, and wish to compare the strength of that signal
across traits. However, analytical tools for such comparisons have
largely remained underdeveloped. Here we evaluate the efficacy of one
commonly used parameter (Pagel's \(\lambda\)) to estimate the strength
of phylogenetic signal in phenotypic traits, and evaluate the degree to
which \(\lambda\) correctly identifies known levels of phylogenetic
signal. We find that \(\lambda\) behaves as a Bernoulli random variable,
and that estimates are increasingly skewed at larger and smaller input
levels of phylogenetic signal. Further, the precision of \(\lambda\) in
estimating actual levels of phylogenetic signal is often inaccurate, and
biological interpretations of the strength of phylogenetic signal based
on \(\lambda\) are therefore compromised. We propose a standardized
effect size based on \(\kappa\), (\(Z_\kappa\)), which measures the
strength of phylogenetic signal more reliably than does \(\lambda\), and
places that signal on a common scale for statistical comparison. We
develop tests based on \(Z_\kappa\) to provide a mechanism for formally
comparing the strength of phylogenetic signal across datasets, in much
the same manner as effect sizes may be used to summarize patterns in
quantitative meta-analysis. Our approach extends the phylogenetic
comparative toolkit to address hypotheses that compare the strength of
phylogenetic signal between various phenotypic traits, even when those
traits are found in different evolutionary lineages or have different
units or scales.
\end{abstract}

@@

Several analytical tools have been developed to quantify phylogenetic
signal in phenotypic datasets (13, 14, 16--19), and their statistical
properties -- namely type I error rates and statistical power -- have
been investigated to determine under what conditions phylogenetic signal
can be detected (15, 18, 20--25). One of the most widely used methods
for characterizing phylogenetic signal is Pagel's \(\lambda\) (13),
which transforms the lengths of the internal branches of the phylogeny
to improve the fit of data to the phylogeny via maximum likelihood (13,
26). When incorporated in PGLS, \(\lambda\) serves as a tuning parameter
(ranging from \(0\rightarrow1\)) which is optimized via log-likelihood
profiling while evaluating the covariation between the dependent and
independent variables, given the phylogeny (13, 26). To infer whether
phylogenetic signal differs from no signal or a Brownian motion model of
evolutionary divergence, the observed model fit using \(\hat\lambda\)
may be statistically compared to that using \(\lambda=0\) or
\(\lambda=1\) via likelihood ratio tests (26--28) or confidence limits
(29). Another widely used measure is Blomberg's \(\kappa\) (14), which
measures the ratio of observed trait variation to the amount expected
under Brownian motion. Blomberg's \(\kappa\) can be treated as a test
statistic by employing a permutation test to generate its sampling
distribution (14, 18) for determining whether significant phylogenetic
signal is present in data. Both \(\lambda\) and \(\kappa\) seem
intuitive to interpret, as a value of \(0\) for both corresponds to no
phylogenetic signal, while a value of \(1\) corresponds to the amount of
phylogenetic signal expected under Brownian motion. Thus, it is tempting
to regard both \(\lambda\) and \(\kappa\) as descriptive statistics that
measure the relative strength of phylogenetic signal, providing an
estimate of its magnitude for comparison.

The appeal of Pagel's \(\lambda\) and Blomberg's \(\kappa\) as
descriptive statistics is that they provide a basis for interpreting
``weak'' versus ``strong'' phylogenetic signal; i.e., small versus large
values of \(\hat{\lambda}\) or \(\kappa\), respectively, in a
comparative sense (30--32). Nonetheless, an important question that has
yet to be considered is whether these statistics are or can be converted
to effect sizes for comparative analyses across datasets? To be
statistics representing phylogenetic signal, they should have reliable
distributional properties, which could be revealed with simulation
experiments. For instance, as a proportional random variable bounded by
\(0\) and \(1\), we might expect that \(\hat{\lambda}\) follows a
Bernoulli distribution (\textbf{add ref}); i.e., branch lengths in a
tree are scaled proportionally to the probability that data arise from a
BM process. Given a known \(\lambda\) value used to generate random data
on a tree, we would also expect that the mean of an empirical sampling
distribution of \(\hat{\lambda}\) would approximately equal \(\lambda\);
the dispersion of \(\hat{\lambda}\) would be largest at intermediate
values of \(\lambda\), \(\hat{\lambda}\) would be predictable over the
range of \(\lambda\) with respect to treesize; the distribution of
\(\hat{\lambda}\) would be symmetric at intermediate values of
\(\lambda\) and more skewed toward values of 0 or 1; and that the
distribution of \(\hat{\lambda}\) will be more platykurtic at
intermediate values of \(\lambda\), becoming more leptokurtic toward 0
and 1 (\textbf{add same ref}). Prior work (20) seems to support some of
these conjectures, based superficially on statistical moments for a
given tree size (mean, variance, skewness, and kurtosis; see Fig. 2 of
(20)). However, because the ``strength of Brownian motion'' was
simulated as a varied weighted-average of data simulated on trees with
\(\lambda=0\) and \(\lambda=1\) and not prescribed values of \(\lambda\)
(20), interpretation of these patterns is challenging.

By contrast, Blomberg's \(\kappa\), which is positively ubounded, might
be expected to follow a normal distribution (\textbf{add ref}). Thus we
might expect that \(\kappa\) is symmetrically distributed across
different strengths of phylogenetic signal, for any \(\lambda\) used to
generate data. This attribute seemed less reasonable based on the
simulations performed by Münkemüller et al.~(20), as distibutions were
postively skewed, suggesting the Blomberg's \(\kappa\) might not behave
as a statistic that follows a noraml distribution. However, because
their simulations used a weighted combination of simulated phylogenetic
signal strengths, strong inferences are not possible (and distributional
attributes were not the intended result of their simulations). For both
Pagel's \(\lambda\) or Blomberg's \(\kappa\), evaluation of statistical
moments across a range of \(\lambda\) used to generate data would be
valuable for adjudicating the relibalilty of these statistics.
Furthermore, these are statistics that appear to have expected values
that vary with tree size (20), making comparisons across studies
challenging. Therefore, transformation of these statistics into
\(Z\)-scores in the same simulation experiments would allow evaluation
of the efficacy of each statistic to yield effect sizes that could be
used for comparative anlyses.

Here we used simulation experiments to compare the distributional
attributes of \(\hat{\lambda}\) and Blomberg's \(\kappa\), plus their
effect sizes (\(Z\)-scores), across a range of tree size and
phylogenetic signal strength. We find that estimates of
\(\hat{\lambda}\) are increasingly skewed at larger and smaller input
levels of phylogenetic signal, vary widely for a given input value of
\(\lambda\), and that the precision of \(\hat{\lambda}\) is not constant
across its range. By contrast, estimates of \(\kappa\) are more
consistent across tree sizes, and are normally distributed across the
range of input levels of \(\lambda\). We then propose an effect size
based on \(\kappa\) (\(Z_{\kappa}\)), which provides consistent
estimates of the strength of phylogenetic signal across tree sizes, and
facilitates quantitative comparisons of the relative strength of
phylogenetic signal across datasets.

@@

\textbf{Lambda (\(\lambda\)) estimates of phylogenetic signal are
innacurate.} Computer simulations revealed that for \(\hat{\lambda}\),
the distributional expectations of a Bernoulli variable were mostly
upheld. First, the mean value of \(\hat{\lambda}\) increased as
\(\lambda\) increased (Fig. 1 black line); the standard deviation of
\(\hat{\lambda}\) was largest at intermediate values of \(\lambda\) and
smallest at extreme values (Fig. 1 red line); and the distributions of
\(\hat{\lambda}\) were increasing skewed at extreme values of
\(\lambda\) but rather normal at intermediate values (Fig. 1 blue line).
For small tree sizes, it was also clear that distributions were more
platykurtik at intermeidate values of \(\hat{\lambda}\). Second,
\(\hat{\lambda}\) was a negatively-biased estimate of \(\lambda\) for
small tree sizes, as the mean value was consistently less than the input
\(\lambda\) value. Third, standard deviations of \(\hat{\lambda}\) were
negatively associated with tree size, and for tress of 128 species or
less, \(\hat{\lambda}\) was quite variable, except for cases when
\(\lambda\) was near or equal to \(1\). As such its statistical
properties make it insuitable as an effect size for measuring the
strength of phylogenetic signal.

@@

\textbf{Kappa (\(\kappa\)) estimates of phylogenetic signal are more
stable.} Results for \(\kappa\) demonstrated better statistical
properties. First, mean values increased consistently with \(\lambda\)
irespective of tree size (Fig. 2 black line); the standard deviation of
\(\kappa\) was consistent across tree sizes (Fig. 2 red line); and
\(\kappa\) was normally distributed regardless of \(\lambda\) or tree
size (Fig. 2 blue line). The standard deviation of \(\kappa\) increased
with \(\lambda\), somewhat tracking the mean, but was always less than
the corresponding means. This is perhaps unsurprising, as \(\kappa\) is
bound by 0, and was never large for small values of \(\lambda\).
Finally, distributions of \(\kappa\) displayed slight skewing only at
small tree sizes and when \(\lambda\) was near or equal to 1. This
differs from the results of (20), where skewing appears to be from
combining random values generated independently.

@@

\textbf{Effect sizes from \(\kappa\) (\(Z_{\kappa}\)) better
characterize phylogenetic signal.} We derived effect sizes (Z-scores)
for both \(\lambda\) and \(\kappa\) (see Methods) and evaluated them
relative to known phylogenetic signal. Both \(Z_{\lambda}\) and
\(Z_{\kappa}\) were associated with input phylogenetic signal
(\(\lambda\)), indicating that both statistics captured the observed
signal (Fig. 3). However, effect sizes from \(\hat{\lambda}\) made
little sense, as they were more strongly associated with tree size than
the actual phylogenetic signal in the data (Fig. 3). By contrast,
\(Z_{\kappa}\) was much more consistent accross tree sizes, and
exhibited a much stronger association with phylogenetic signal strength
as compared to tree size (Fig. 3). Thus between the two statistics,
\(Z_{\kappa}\) is a more reliable measure of the strength of
phylogenetic signal, and may be used to compare levels of phylogenetic
signal across datasets.

@@

\textbf{A test statistic (\(\hat{Z}_{12}\)) allows meaningful
comparisons across datasets.} We derived a two-sample test statistic
(\(\hat{Z}_{12}\)) for statistically comparing the strength of
phylogenetic signal across datasets (see Methods). To demonstrate its
utility, we compared \(Z_{\kappa}\) for two ecologically-relevant traits
in plethodontid salamander (Fig. 4): surface area to volume ratios
(SA:V) and relative body width (\(\frac{BW}{SVL}\)) (33, 34). While both
contained significant phylogenetic signal, tests based on
\(\hat{Z}_{12}\) revealed that the degree of phylogenetic signal was
significantly stronger in SA:V (Fig. 4). Biologically, this observation
corresponds with the fact that tropical species -- which form a
monophyletic group within plethodontids -- display greater variation in
SA:V which covaries with disparity in their climatic niches (34).

@@

where \(\kappa_{obs}\) is the observed phylogenetic signal, and
\(\hat\mu_\kappa\) and \(\hat\sigma_\kappa\) are the mean and standard
deviation of the empirical sampling distribution of \(\kappa\) obtained
via permutation. The empirical sampling distribution of \(\kappa\) is
first transformed via Box-Cox to better adhere to the assumption of
normality. \hfill\break

For \(\lambda\) deriving an effect size is more challenging, it does not
have a sampling distribution from which the standard error (and thus
confidence intervals) may be obtained. Thus, its confidence intervals
are generated for the values of \(\lambda\) that intersect the
log-likehihood profile for corresponding percentiles of the \(\chi^2\)
distribution used to compare the compare the putative appropriate model
to a null model with \(\lambda = 0\) {[}\textbf{add ref: MLC thinks
Boettiger paper?}{]}. As such, an effect size for \(\lambda\) may be
found as:

@@

where, \(\hat{\lambda}\) is the maximized likelihood value of
\(\lambda\) and \(\chi^{2}_{\hat{\lambda}}\) is the likelihood ratio
statistic for the value. Note that alternative formulations could be
envisioned: \(Z_{\lambda} = d \sqrt{\chi^2_{\hat{\lambda}}}\), where
\(d\) is a binary value \((-1,1)\) to indicate a direction based on
whether \(\hat{\lambda}\) is below or above a critical value of
\(\lambda\), for a quantile from a \(\chi^{2}\) distribution at a
probability of 0.5. However, preliminary investigations found that
mapping \(\chi^2_{\hat{\lambda}}\) values in this manner did not produce
effect sizes that were symmetrical about \(Z = 0\), as the mapping was
not linear and the log-likelihood profiles can be rather flat for small
trees.

@@

\textbf{Empirical Data.} Surface area to volume ratios (SA:V) and
relative body width (\(\frac{BW}{SVL}\)) measures were obtained from
individuals of 305 species (2,781 and 3,371 respectively), from which
species means were obtained (33, 34). A time-dated molecular phylogeny
for the group (47) was then pruned to match the species in the
phenotypic dataset. The phylogenetic signal in each trait was then
characterized using \(\kappa\), which was converted to its effect size
(\(Z_\kappa\)) using \texttt{geomorph} 3.3.1 (48, 49).

>>>>

\author[a,1]{Dean C. Adams}
\author[a,b]{Erica K. Baken}
\author[b]{Michael L. Collyer}

  \affil[a]{Department of Ecology, Evolution, and Organismal Biology, Iowa State
University, Ames, Iowa, 50010. USA.}
  \affil[b]{Department of Science, Chatham University, Pittsburgh, Pennsylvania,
15232. USA.}

@@

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Evolutionary biologists wish to quantify and compare the strength of
phylogenetic signal across traits, but analytical tools for these
comparisons are generally lacking. Here we develop a standardized effect
size based on \(\kappa\), (\(Z_\kappa\)), which measures the strength of
phylogenetic signal on a common statistical scale, and provides a
mechanism for formally comparing the strength of phylogenetic signal
across datasets. Additionally, we find that a commonly used parameter
(Pagel's \(\lambda\)) is unsuitable for this purpose. Our procedure
enables biologists to quantitatively address hypotheses that compare the
strength of phylogenetic signal between various phenotypic traits, even
when those traits are found in different evolutionary lineages or have
different units or scales.}

@@

\authordeclaration{The authors declare no conflict of interest. \hfill\break

Data deposition: Data for the empirical example may be found on DRYAD:
\url{doi:10.5061/dryad.b554m44} and \url{doi:10.5061/dryad.59zw3r23m}.
R-scripts for simulation tests are found on Github: XXX. Computer code
for implementing the two-sample comparison of effect sizes is found in
geomorph:
\url{https://cran.r-project.org/web/packages/geomorph/index.html}
\hfill\break

\textsuperscript{1}To whom correspondence should be addressed. E-mail:
\href{mailto:dcadams@iastate.edu}{\nolinkurl{dcadams@iastate.edu}}}

@@

\begin{abstract}
Macroevolutionary studies frequently characterize the phylogenetic
signal in phenotypes, and wish to compare the strength of that signal
across traits. However, analytical tools for such comparisons have
largely remained underdeveloped. Here we evaluate the efficacy of one
commonly used parameter (Pagel's \(\lambda\)) to estimate the strength
of phylogenetic signal in phenotypic traits, and evaluate the degree to
which \(\lambda\) correctly identifies known levels of phylogenetic
signal. We find that \(\lambda\) behaves as a Bernoulli random variable,
and that estimates are increasingly skewed at larger and smaller input
levels of phylogenetic signal. Further, the precision of \(\lambda\) in
estimating actual levels of phylogenetic signal is often inaccurate, and
biological interpretations of the strength of phylogenetic signal based
on \(\lambda\) are therefore compromised. As an alternative, we propose
a standardized effect size based on \(\kappa\), (\(Z_\kappa\)), which
measures the strength of phylogenetic signal more reliably than does
\(\lambda\), and places that signal on a common scale for statistical
comparison. We develop tests based on \(Z_\kappa\) to provide a
mechanism for formally comparing the strength of phylogenetic signal
across datasets, in much the same manner as effect sizes may be used to
summarize patterns in quantitative meta-analysis. Our approach extends
the phylogenetic comparative toolkit to address hypotheses that compare
the strength of phylogenetic signal between various phenotypic traits,
even when those traits are found in different evolutionary lineages or
have different units or scales.
\end{abstract}

@@

Several analytical tools have been developed to quantify phylogenetic
signal in phenotypic datasets (13, 14, 16--19), and their statistical
properties -- namely type I error rates and statistical power -- have
been investigated to determine under what conditions phylogenetic signal
can be detected (15, 18, 20--25). One of the most widely used methods
for characterizing phylogenetic signal is Pagel's \(\lambda\) (13),
which transforms the lengths of the internal branches of the phylogeny
to improve the fit of data to the phylogeny via maximum likelihood (13,
26). When incorporated in PGLS, \(\lambda\) serves as a tuning parameter
which is optimized via log-likelihood profiling while evaluating the
covariation between the dependent and independent variables, given the
phylogeny (13, 26). To infer whether phylogenetic signal differs from no
signal or a Brownian motion model of evolutionary divergence, the
observed model fit using \(\hat\lambda\) may be statistically compared
to that using \(\lambda=0\) or \(\lambda=1\) via likelihood ratio tests
(26--28) or confidence limits (29). Another widely used measure is
Blomberg's \(\kappa\) (14), which measures the ratio of observed trait
variation to the amount expected under Brownian motion. Blomberg's
\(\kappa\) can be treated as a test statistic by employing a permutation
test to generate its sampling distribution (14, 18) for determining
whether significant phylogenetic signal is present in data. Both
\(\lambda\) and \(\kappa\) seem intuitive to interpret, as a value of
\(0\) for both corresponds to no phylogenetic signal, while a value of
\(1\) corresponds to the amount of phylogenetic signal expected under
Brownian motion. Thus, it is tempting to regard both \(\lambda\) and
\(\kappa\) as descriptive statistics that measure the relative strength
of phylogenetic signal, providing an estimate of its magnitude for
comparison.

The appeal of Pagel's \(\lambda\) and Blomberg's \(\kappa\) as
descriptive statistics is that they provide a basis for interpreting
``weak'' versus ``strong'' phylogenetic signal; i.e., small versus large
values of \(\hat{\lambda}\) or \(\kappa\), respectively, in a
comparative sense (30--32). Nonetheless, an important question that has
yet to be considered is whether these statistics are or can be converted
to effect sizes for comparative analyses across datasets? To be
statistics representing phylogenetic signal, they should have reliable
distributional properties, which could be revealed with simulation
experiments. For instance, as a proportional random variable bounded by
\(0\) and \(1\), we might expect that \(\hat{\lambda}\) follows a
Bernoulli distribution (\textbf{add ref}); i.e., branch lengths in a
tree are scaled proportionally to the probability that data arise from a
BM process. Given a known \(\lambda\) value used to generate random data
on a tree, we would also expect that the mean of an empirical sampling
distribution of \(\hat{\lambda}\) would approximately equal \(\lambda\);
the dispersion of \(\hat{\lambda}\) would be largest at intermediate
values of \(\lambda\), \(\hat{\lambda}\) would be predictable over the
range of \(\lambda\) with respect to treesize; the distribution of
\(\hat{\lambda}\) would be symmetric at intermediate values of
\(\lambda\) and more skewed toward values of 0 or 1; and that the
distribution of \(\hat{\lambda}\) will be more platykurtic at
intermediate values of \(\lambda\), becoming more leptokurtic toward 0
and 1 (\textbf{add same ref}). Prior work (20) seems to support some of
these conjectures, based superficially on statistical moments for a
given tree size (mean, variance, skewness, and kurtosis; see Fig. 2 of
ref. (20)). However, because the ``strength of Brownian motion'' was
simulated as a varied weighted-average of data simulated on trees with
\(\lambda=0\) and \(\lambda=1\) and not as prescribed values of
\(\lambda\) (20), interpretation of these patterns is challenging.

By contrast, Blomberg's \(\kappa\), which is positively ubounded, might
be expected to follow a normal distribution (\textbf{add ref}). Thus we
might expect that \(\kappa\) is symmetrically distributed across
different strengths of phylogenetic signal, for any \(\lambda\) used to
generate data. This attribute seemed less reasonable based on the
simulations performed by Münkemüller et al.~(20), which suggested that
distributions were positively skewed and that Blomberg's \(\kappa\)
might not behave as a statistic that follows a normal distribution.
However, because their simulations used a weighted combination of
simulated phylogenetic signal strengths, strong inferences are not
possible (and distributional attributes were not the intended result of
their simulations). Thus, for both Pagel's \(\lambda\) or Blomberg's
\(\kappa\), evaluation of statistical moments across a range of
\(\lambda\) used to generate data would be valuable for adjudicating the
relibalilty of these statistics. Furthermore, these are statistics that
appear to have expected values that vary with tree size (20), making
comparisons across studies challenging. Therefore, transformation of
these statistics into \(Z\)-scores in the same simulation experiments
would allow evaluation of the efficacy of each statistic to yield effect
sizes that could be used for comparisons of the strength of phylogenetic
signal across traits and clades.

Here we used simulation experiments to compare the distributional
attributes of \(\hat{\lambda}\) and \(\kappa\), plus their effect sizes
(\(Z\)-scores), across a range of tree size and phylogenetic signal
strength. We find that estimates of \(\hat{\lambda}\) are increasingly
skewed at larger and smaller input levels of phylogenetic signal and at
smaller tree sizes, vary widely for a given input value of \(\lambda\),
and that the precision of \(\hat{\lambda}\) is not constant across its
range. By contrast, estimates of \(\kappa\) are more consistent across
tree sizes, and are normally distributed across the range of input
levels of \(\lambda\), making \(\kappa\) a more reliable statistic. We
then propose an effect size based on \(\kappa\), (\(Z_{\kappa}\)), which
provides consistent estimates of the strength of phylogenetic signal
across tree sizes and signal strength, and facilitates quantitative
comparisons of the relative strength of phylogenetic signal across
datasets.

@@

\textbf{Lambda (\(\lambda\)) estimates of phylogenetic signal are
innacurate.} Computer simulations revealed that for \(\hat{\lambda}\),
the distributional expectations of a Bernoulli variable were mostly
upheld. First, the mean value of \(\hat{\lambda}\) did increase as
\(\lambda\) increased, but it was negatively-biased (particularly for
small tree sizes), and was consistently less than the input \(\lambda\)
value across most of its range (Fig. 1 black line). Second, the standard
deviation of \(\hat{\lambda}\) was largest at intermediate values of
\(\lambda\) and smallest at extreme values, implying that the precision
in estimating \(\lambda\) varied across the range of input values (Fig.
1 red line). Additionally, standard deviations of \(\hat{\lambda}\) were
negatively associated with tree size, and for tress of 128 species or
less, \(\hat{\lambda}\) was quite variable, except for cases when
\(\lambda\) was near or equal to \(1\). Third, the distributions of
\(\hat{\lambda}\) were not normal across its range, but became
increasingly skewed at more extreme values of \(\lambda\) (Fig. 1 blue
line). For small tree sizes, it was also clear that distributions were
more platykurtic at intermeidate values of \(\hat{\lambda}\). Taken
together these results reveal that \(\hat\lambda\) inconsistently
estimated phylogenetic signal, both across tree sizes and across the
range of input values. Additional simulations (Supplemental Information)
revealed that incorporating \(\lambda\) in PGLS anova and regression did
not adversely affect the statistical properties of model fitting (type I
error, power, parameter estimation); thus, it is reasonable to include
\(\lambda\) in PGLS as a parameter for tuning the degree of phylogenetic
signal in the dependent variables during the analysis. However, the
statistical properties revealed in Fig. 1 demonstrate that \(\lambda\)
is unsuitable as an effect size for measuring the strength of
phylogenetic signal in data.

@@

\textbf{Kappa (\(\kappa\)) estimates of phylogenetic signal are more
stable.} Simulation results for \(\kappa\) demonstrated that this
measure displayed better statistical properties. First, as expected,
mean values of \(\kappa\) increased consistently with increasing signal
(\(\lambda\)) irrespective of tree size (Fig. 2 black line).
Additionally, the standard deviation of \(\kappa\) was consistent across
tree sizes (Fig. 2 red line), and while it increased with \(\lambda\),
it was always less than the corresponding mean. This finding is perhaps
unsurprising, as \(\kappa\) is bound by 0, and was never large for small
values of \(\lambda\). Importantly, \(\kappa\) was normally distributed
across the range of input \(\lambda\), and remained consistent in this
pattern regardless of tree size (Fig. 2 blue line). In fact,
distributions of \(\kappa\) displayed slight skewing only at small tree
sizes and when \(\lambda\) was near or equal to \(1\). This result
differs from those of (20), where skewing appears to be from combining
random values generated independently. These findings reveal that
\(\kappa\) is more reliable as an estimate of phylogenetic signal.

@@

\textbf{Effect sizes from \(\kappa\) (\(Z_{\kappa}\)) better
characterize phylogenetic signal.} To measure the strength of
phylogenetic signal on a common scale, we derived effect sizes
(Z-scores) for both \(\lambda\) and \(\kappa\) (see Methods) and
evaluated them relative to known phylogenetic signal. Both
\(Z_{\lambda}\) and \(Z_{\kappa}\) were associated with input
phylogenetic signal (\(\lambda\)), indicating that both statistics
captured the observed signal (Fig. 3). However, effect sizes from
\(\hat{\lambda}\) made little sense, as they were more strongly
associated with tree size than they were with the actual phylogenetic
signal in the data (Fig. 3). By contrast, \(Z_{\kappa}\) was much more
consistent accross tree sizes, and exhibited a much stronger association
with phylogenetic signal strength as compared to tree size (Fig. 3).
Thus between the two statistics, \(Z_{\kappa}\) is a more reliable
measure of the strength of phylogenetic signal, and may be used to
compare levels of phylogenetic signal across datasets.

@@

\textbf{A test statistic (\(\hat{Z}_{12}\)) allows meaningful
comparisons across datasets.} To statistically compare the strength of
phylogenetic signal across datasets we derived a two-sample test
statistic (\(\hat{Z}_{12}\): see Methods). As an example of its utility,
we compared \(Z_{\kappa}\) for two ecologically-relevant traits in
plethodontid salamander (Fig. 4): surface area to volume ratios (SA:V)
and relative body width (\(\frac{BW}{SVL}\)) (33, 34). While both traits
contained significant phylogenetic signal, tests based on
\(\hat{Z}_{12}\) revealed that the degree of phylogenetic signal was
significantly stronger in SA:V (Fig. 4). Biologically, this observation
may be interpreted by the fact that tropical species -- which form a
monophyletic group within plethodontids -- display greater variation in
SA:V, which covaries with disparity in their climatic niches (34). Thus,
greater phylogenetic signal in SA:V is to be expected.

@@

where \(\kappa_{obs}\) is the observed phylogenetic signal, and
\(\hat\mu_\kappa\) and \(\hat\sigma_\kappa\) are the mean and standard
deviation of the empirical sampling distribution of \(\kappa\) obtained
via permutation. The empirical sampling distribution of \(\kappa\) is
first transformed via Box-Cox to better adhere to the assumption of
normality.

For \(\lambda\), deriving an effect size is more challenging, as
\(\lambda\) does not have a sampling distribution from which the
standard error (and thus confidence intervals) may be obtained.
Confidence intervals are tehrefore generated for the values of
\(\lambda\) that intersect the log-likehihood profile for corresponding
percentiles of the \(\chi^2\) distribution used to compare the putative
model to a null model with \(\lambda = 0\) {[}\textbf{add ref: MLC
thinks Boettiger paper?}{]}. Thus, an effect size for \(\lambda\) may be
found as:

@@

where, \(\hat{\lambda}\) is the maximized likelihood value of
\(\lambda\) and \(\chi^{2}_{\hat{\lambda}}\) is the likelihood ratio
statistic for the value. Note that alternative formulations could be
envisioned: \(Z_{\lambda} = d \sqrt{\chi^2_{\hat{\lambda}}}\), where
\(d\) is a binary value \((-1,1)\) to indicate a direction based on
whether \(\hat{\lambda}\) is below or above a critical value of
\(\lambda\), for a quantile from a \(\chi^{2}\) distribution at a
probability of 0.5. However, preliminary investigations found that
mapping \(\chi^2_{\hat{\lambda}}\) values in this manner did not produce
effect sizes that were symmetrical about \(Z = 0\), as the mapping was
not linear and the log-likelihood profiles can be rather flat for small
trees. \textbf{Is alternative needed here?}

@@

\textbf{Empirical Data.} Surface area to volume ratios (SA:V) and
relative body width (\(\frac{BW}{SVL}\)) measures were obtained from as
species means from individuals of 305 species, from which species means
were obtained (33, 34). A time-dated molecular phylogeny for the group
(47) was pruned to match the species in the phenotypic dataset. The
phylogenetic signal in each trait was then characterized using
\(\kappa\), which was converted to its effect size (\(Z_\kappa\)) using
\texttt{geomorph} 3.3.1 (48, 49), and routines by the authors
(\textbf{to be incorporated in \texttt{geomorph} upon manuscript
acceptance}).
